### 2006-05-11 paper/tags/arne_diss_submit/systemfit_code.tex
created tag for the revision that is included in Arnes dissertation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@667 591db2bd-e0a0-4d22-8c19-2192dbb7da31

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection[Basic function systemfit]{The basic function \code{systemfit}}
The basic functionality of this package is provided by the
function \code{systemfit}.
This function estimates the equation system as described
in sections~\ref{sec:Estimation}.
If parameter restrictions are provided, the formulas in
section~\ref{sec:Restrictions} are applied.
Furthermore, the user can control several details of the estimation.
For instance, the
formula to calculate the residual covariance matrix
(see section~\ref{sec:residcov}),
the degrees of freedom for the t-tests
(see section~\ref{sec:degreesOfFreedom}), or
the formula for the 3SLS estimation
(see sections~\ref{sec:Estimation} and~\ref{sec:Restrictions})
can be specified by the user.
The \code{systemfit} function returns many objects
that users might be interest in.
A complete list is available in the documentation of this function
that is included in the package.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Statistical tests}
The statistical tests described in
sections~\ref{sec:testingRestrictions} and~\ref{sec:hausman}
are implemented as specified in these sections.
The functions \code{ftest.systemfit}, \code{waldtest.systemfit} and
\code{lrtest.systemfit} test linear restrictions on the
estimated parameters.
On the other hand, the function \code{hausman.systemfit}
tests the consistency of the 3SLS estimator.
All functions return the empirical test statistic,
the degree(s) of freedom, and the p-value.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiency of the code}
We have followed \cite{bates04} to make the code faster and more stable.
First, if a formula contains an inverse of a matrix
that is post-multiplied by a vector,
we use \code{solve(A, b)} instead of \code{solve(A) \%*\% b}.
Second, we calculate crossproducts
by \code{crossprod(X)} or \code{crossprod(X, y)}
instead of \code{t(X) \%*\% X} or \code{t(X) \%*\% y},
respectively.

The matrix $\Omega^{-1}$ that is used to compute the estimated
coefficients and their covariance matrix is of size
$( G \cdot T ) \times ( G \cdot T )$
(see sections~\ref{sec:Estimation} and~\ref{sec:Restrictions}).
In case of large data sets, this matrix $\Omega^{-1}$ gets really huge
and needs a lot of memory.
Therefore, we use the following transformation and compute $X' \Omega^{-1}$
by deviding the $X$ matrix into submatrices,
doing some calculations with these submatrices,
adding up some of these submatrices, and
finally putting the submatrices together:
\begin{equation}
X' \Omega^{-1}
%= X' \left( \Sigma^{-1} \otimes I \right)
= \sum_{i=1} \left[ \begin{array}{c}
   \sigma^{1i} {X^1} \\
   \sigma^{2i} {X^2} \\
   \vdots \\
   \sigma^{Gi} {X^G} \\
\end{array} \right]'
\end{equation}
where $\sigma^{ij}$ are the elements of the matrix $\Sigma^{-1}$,
and $X^i$ is a submatrix of $X$ that contains the rows
that belong to the $i$'s equation.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection[Basic function systemfit]{The basic function \code{systemfit}}
The basic functionality of this package is provided by the
function \code{systemfit}.
This function estimates the equation system as described
in sections~\ref{sec:Estimation}.
If parameter restrictions are provided, the formulas in
section~\ref{sec:Restrictions} are applied.
Furthermore, the user can control several details of the estimation.
For instance, the
formula to calculate the residual covariance matrix
(see section~\ref{sec:residcov}),
the degrees of freedom for the $t$ tests
(see section~\ref{sec:degreesOfFreedom}), or
the formula for the 3SLS estimation
(see sections~\ref{sec:Estimation} and~\ref{sec:Restrictions})
can be specified by the user.
The \code{systemfit} function returns many objects
that users might be interest in.
A complete list is available in the documentation of this function
that is included in the package.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Statistical tests}
The statistical tests described in
sections~\ref{sec:testingRestrictions} and~\ref{sec:hausman}
are implemented as specified in these sections.
The functions \code{ftest.systemfit}, \code{waldtest.systemfit} and
\code{lrtest.systemfit} test linear restrictions on the
estimated parameters.
On the other hand, the function \code{hausman.systemfit}
tests the consistency of the 3SLS estimator.
All functions return the empirical test statistic,
the degree(s) of freedom, and the $p$ value.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiency of the code}
We have followed \cite{bates04} to make the code faster and more stable.
First, if a formula contains an inverse of a matrix
that is post-multiplied by a vector,
we use \code{solve(A,b)} instead of \code{solve(A) \%*\% b}.
Second, we calculate crossproducts
by \code{crossprod(X)} or \code{crossprod(X,y)}
instead of \code{t(X) \%*\% X} or \code{t(X) \%*\% y},
respectively.

The matrix $\Omega^{-1}$ that is used to compute the estimated
coefficients and their covariance matrix is of size
$( G \cdot T ) \times ( G \cdot T )$
(see sections~\ref{sec:Estimation} and~\ref{sec:Restrictions}).
In case of large data sets, this matrix $\Omega^{-1}$ gets really huge
and needs a lot of memory.
Therefore, we use the following transformation and compute $X' \Omega^{-1}$
by dividing the $X$ matrix into submatrices,
doing some calculations with these submatrices,
adding up some of these submatrices, and
finally putting the submatrices together:
\begin{equation}
X' \Omega^{-1}
%= X' \left( \Sigma^{-1} \otimes I \right)
= \sum_{i=1} \left[ \begin{array}{c}
   \sigma^{1i} {X^1} \\
   \sigma^{2i} {X^2} \\
   \vdots \\
   \sigma^{Gi} {X^G} \\
\end{array} \right]'
\end{equation}
where $\sigma^{ij}$ are the elements of the matrix $\Sigma^{-1}$,
and $X^i$ is a submatrix of $X$ that contains the rows
that belong to the $i$'s equation.

### 2006-05-11 paper/tags/arne_diss_submit/systemfit_intro.tex
created tag for the revision that is included in Arnes dissertation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@667 591db2bd-e0a0-4d22-8c19-2192dbb7da31

Many theoretical models that are econometrically estimated
consist of more than one equation.
The disturbance terms of these equations are likely
to be contemporaneously correlated,
because unconsidered factors
that influence the disturbance term in one equation
probably influence the disturbance terms in other equations.
Ignoring this contemporaneous correlation
and estimating these equations separately
leads to inefficient parameter estimates.
However, estimating all equations simultaneously,
taking the covariance structure of the residuals into account,
leads to efficient estimates.
This estimation procedure is generally called
"Seemingly Unrelated Regression" (SUR) \citep{zellner62}.
Another reason to estimate an equation system simultaneously are
cross-equation parameter restrictions.%
\footnote{
Especially the economic theory suggests many cross-equation parameter
restrictions (e.g.\ the symmetry restriction in demand models).
}
These restrictions can be tested and/or imposed only in a simultaneous
estimation approach.

Furthermore, these models can contain variables
that appear on the left-hand side in one equation
and on the right-hand side of another equation.
Ignoring the endogeneity of these variables can lead to inconsistent
parameter estimates.
This simultaneity bias can be corrected for in each equation by
applying a "Two-Stage Least Squares" (2SLS) method or for all
equations simultaneously when combined with SUR resulting in a
"Three-Stage Least Squares" (3SLS) estimation of the system of
equations.

>>>>

Many theoretical models that are econometrically estimated
consist of more than one equation.
The disturbance terms of these equations are likely
to be contemporaneously correlated,
because unconsidered factors
that influence the disturbance term in one equation
probably influence the disturbance terms in other equations.
Ignoring this contemporaneous correlation
and estimating these equations separately
leads to inefficient parameter estimates.
However, estimating all equations simultaneously,
taking the covariance structure of the residuals into account,
leads to efficient estimates.
This estimation procedure is generally called
°Seemingly Unrelated Regression° (SUR) \citep{zellner62}.
Another reason to estimate an equation system simultaneously are
cross-equation parameter restrictions.%
\footnote{
Especially the economic theory suggests many cross-equation parameter
restrictions (e.g.\ the symmetry restriction in demand models).
}
These restrictions can be tested and/or imposed only in a simultaneous
estimation approach.

Furthermore, these models can contain variables
that appear on the left-hand side in one equation
and on the right-hand side of another equation.
Ignoring the endogeneity of these variables can lead to inconsistent
parameter estimates.
This simultaneity bias can be corrected for in each equation by
applying a °Two-Stage Least Squares° (2SLS) method or for all
equations simultaneously when combined with SUR resulting in a
°Three-Stage Least Squares° (3SLS) estimation of the system of
equations.

### 2006-05-11 paper/tags/arne_diss_submit/systemfit_other.tex
created tag for the revision that is included in Arnes dissertation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@667 591db2bd-e0a0-4d22-8c19-2192dbb7da31

The WLS, SUR, W2SLS and 3SLS parameter estimates are consistent,
if the estimated residual covariance matrix is calculated
using the residuals from a first-step OLS or 2SLS estimation.
There exists also an alternative slightly different approach.%
\footnote{
For instance, this approach is applied by
the command °TSCS° of the software LIMDEP that carries out SUR estimations
in which all coefficient vectors are constrained to be equal
\citep{greene06}.
}
This alternative approach uses the residuals of a first-step OLS or 2SLS estimation
to apply a WLS or W2SLS estimation on a second step.
Then, it calculates the residual covariance matrix
from the residuals of the second-step estimation
to estimates the model by SUR or 3SLS in a third step.
If no cross-equation restrictions are imposed,
the parameter estimates of OLS and WLS as well as 2SLS and W2SLS are identical.
Hence, in this case both approaches generate the same results.

@@

To our knowledge the question about how to determine the degrees
of freedom for single-parameter t-tests is not comprehensively
discussed in the literature.
While sometimes the degrees of freedom of the entire system
(total number of observations in all equations minus
total number of estimated parameters)
are applied,
in other cases the degrees of freedom of each single equation
(number of observations in the equations minus
number of estimated parameters in the equation)
are used.
Asymptotically, this distinction doesn't make a difference.
However, in many empirical applications, the number of observations
of each equation is rather small, and
therefore, it matters.

If a system of equations is estimated by an unrestricted OLS and
the covariance matrix of the parameters is calculated
by~(\ref{eq:olsCovSingleSigma}),
the estimated parameters and their standard errors are identical
to an equation-wise OLS estimation.
In this case, it is reasonable to use the degrees of freedom of
each single equation,
because this yields the same p-values as the equation-wise
OLS estimation.

In contrast, if a system of equations is estimated with many
cross-equation restrictions and
the covariance matrix of an OLS estimation is calculated
by~(\ref{eq:olsCovSameSigma}),
the system estimation is similar to a single equation estimation.
Therefore, in this case, it seems to be reasonable to use the degrees
of freedom of whole system.

@@

Linear restrictions can be tested by an F test, Wald test or
likelihood-ratio (LR) test.

The F-statistic for systems of equations is
\begin{equation}
F = \frac{
   ( R \hat{\beta} - q )'
   ( R ( X' ( \hat{\Sigma} \otimes I )^{-1} X )^{-1} R' )^{-1}
   ( R \hat{\beta} - q ) /
   j
}{
   \hat{u}' ( \Sigma \otimes I )^{-1} \hat{u} /
   ( M \cdot T - K )
}
\end{equation}
where $j$ is the number of restrictions,
$M$ is the number of equations,
$T$ is the number of observations per equation,
$K$ is the total number of estimated coefficients, and
$\hat{\Sigma}$ is the estimated residual covariance matrix
used in the estimation.
Under the null hypothesis, $F$ has an F-distribution
with $j$ and $M \cdot T - K$ degrees of freedom
\citep[p.\ 314]{theil71}.

The Wald-statistic for systems of equations is
\begin{equation}
W =
   ( R \hat{\beta} - q )'
   ( R \widehat{Cov} [ \hat{\beta} ] R' )^{-1}
   ( R \hat{\beta} - q )
\end{equation}
Asymptotically, $W$ has a $\chi^2$
distribution with $j$ degrees of freedom
under the null hypothesis
\citep[p.\ 347]{greene03}.

The LR-statistic for systems of equations is
\begin{equation}
LR = T \cdot \left(
   log \left| \hat{ \hat{ \Sigma } }_r \right|
   - log \left| \hat{ \hat{ \Sigma } }_u \right|
   \right)
\end{equation}
where $T$ is the number of observations per equation, and 
$\hat{\hat{\Sigma}}_r$ and $\hat{\hat{\Sigma}}_u$ are
the residual covariance matrices calculated by formula (\ref{eq:rcov0})
of the restricted and unrestricted estimation, respectively.
Asymptotically, $LR$ has a $\chi^2$
distribution with $j$ degrees of freedom
under the null hypothesis
\citep[p.\ 349]{greene03}.

@@

\citet{hausman78} developed a test for misspecification.
The null hypotheses of the test is that all exogenous variables are
uncorrelated with all disturbance terms.
Under this hypothesis both the 2SLS and the 3SLS estimator are consistent
but only the 3SLS estimator is (asymptotically) efficient.
Under the alternative hypothesis the 2SLS estimator is consistent
but the 3SLS estimator is inconsistent.
The Hausman test statistic is,
\begin{equation}
  m = \left( \hat{\beta}_{2SLS} - \hat{\beta}_{3SLS} \right)^{'}
      \left( \Cov \left[ \hat{\beta}_{2SLS} \right] -
             \Cov \left[ \hat{\beta}_{3SLS} \right] \right)
      \left( \hat{\beta}_{2SLS} - \hat{\beta}_{3SLS} \right)
\label{eq:hausman}
\end{equation}
where $\hat{\beta}_{2SLS}$ and $\Cov \left[ \hat{\beta}_{2SLS} \right]$ are the estimated
coefficient and covariance matrix from 2SLS estimation, and
$\hat{\beta}_{3SLS}$ and $\Cov \left[ \hat{\beta}_{3SLS} \right]$ are the estimated
coefficients and covariance matrix from 3SLS estimation.
Under the null hypotheses this test statistic has a
$\chi^2$ distribution with degrees of freedom equal to the number of
estimated parameters.

>>>>

The WLS, SUR, W2SLS and 3SLS parameter estimates are consistent
if the estimated residual covariance matrix is calculated
using the residuals from a first-step OLS or 2SLS estimation.
There exists also an alternative slightly different approach.%
\footnote{
For instance, this approach is applied by
the command °TSCS° of the software LIMDEP that carries out SUR estimations
in which all coefficient vectors are constrained to be equal
\citep{greene06}.
}
This alternative approach uses the residuals of a first-step OLS or 2SLS estimation
to apply a WLS or W2SLS estimation on a second step.
Then, it calculates the residual covariance matrix
from the residuals of the second-step estimation
to estimates the model by SUR or 3SLS in a third step.
If no cross-equation restrictions are imposed,
the parameter estimates of OLS and WLS as well as 2SLS and W2SLS are identical.
Hence, in this case both approaches generate the same results.

@@

To our knowledge the question about how to determine the degrees
of freedom for single-parameter $t$ tests is not comprehensively
discussed in the literature.
While sometimes the degrees of freedom of the entire system
(total number of observations in all equations minus
total number of estimated parameters)
are applied,
in other cases the degrees of freedom of each single equation
(number of observations in the equations minus
number of estimated parameters in the equation)
are used.
Asymptotically, this distinction does not make a difference.
However, in many empirical applications, the number of observations
of each equation is rather small, and
therefore, it matters.

If a system of equations is estimated by an unrestricted OLS and
the covariance matrix of the parameters is calculated
by~(\ref{eq:olsCovSingleSigma}),
the estimated parameters and their standard errors are identical
to an equation-wise OLS estimation.
In this case, it is reasonable to use the degrees of freedom of
each single equation,
because this yields the same $p$ values as the equation-wise
OLS estimation.

In contrast, if a system of equations is estimated with many
cross-equation restrictions and
the covariance matrix of an OLS estimation is calculated
by~(\ref{eq:olsCovSameSigma}),
the system estimation is similar to a single equation estimation.
Therefore, in this case, it seems to be reasonable to use the degrees
of freedom of the entire system.

@@

Linear restrictions can be tested by an $F$ test, Wald test or
likelihood ratio (LR) test.

The $F$ statistic for systems of equations is
\begin{equation}
F = \frac{
   ( R \hat{\beta} - q )'
   ( R ( X' ( \hat{\Sigma} \otimes I )^{-1} X )^{-1} R' )^{-1}
   ( R \hat{\beta} - q ) /
   j
}{
   \hat{u}' ( \Sigma \otimes I )^{-1} \hat{u} /
   ( M \cdot T - K )
}
\end{equation}
where $j$ is the number of restrictions,
$M$ is the number of equations,
$T$ is the number of observations per equation,
$K$ is the total number of estimated coefficients, and
$\hat{\Sigma}$ is the estimated residual covariance matrix
used in the estimation.
Under the null hypothesis, $F$ has an $F$ distribution
with $j$ and $M \cdot T - K$ degrees of freedom
\citep[p.\ 314]{theil71}.

The Wald statistic for systems of equations is
\begin{equation}
W =
   ( R \hat{\beta} - q )'
   ( R \widehat{Cov} [ \hat{\beta} ] R' )^{-1}
   ( R \hat{\beta} - q )
\end{equation}
Asymptotically, $W$ has a $\chi^2$
distribution with $j$ degrees of freedom
under the null hypothesis
\citep[p.\ 347]{greene03}.

The LR statistic for systems of equations is
\begin{equation}
LR = T \cdot \left(
   log \left| \hat{ \hat{ \Sigma } }_r \right|
   - log \left| \hat{ \hat{ \Sigma } }_u \right|
   \right)
\end{equation}
where $T$ is the number of observations per equation, and 
$\hat{\hat{\Sigma}}_r$ and $\hat{\hat{\Sigma}}_u$ are
the residual covariance matrices calculated by formula (\ref{eq:rcov0})
of the restricted and unrestricted estimation, respectively.
Asymptotically, $LR$ has a $\chi^2$
distribution with $j$ degrees of freedom
under the null hypothesis
\citep[p.\ 349]{greene03}.

@@

\citet{hausman78} developed a test for misspecification.
The null hypothesis of the test is that all exogenous variables are
uncorrelated with all disturbance terms.
Under this hypothesis both the 2SLS and the 3SLS estimator are consistent
but only the 3SLS estimator is (asymptotically) efficient.
Under the alternative hypothesis the 2SLS estimator is consistent
but the 3SLS estimator is inconsistent.
The Hausman test statistic is,
\begin{equation}
  m = \left( \hat{\beta}_{2SLS} - \hat{\beta}_{3SLS} \right)^{'}
      \left( \Cov \left[ \hat{\beta}_{2SLS} \right] -
             \Cov \left[ \hat{\beta}_{3SLS} \right] \right)
      \left( \hat{\beta}_{2SLS} - \hat{\beta}_{3SLS} \right)
\label{eq:hausman}
\end{equation}
where $\hat{\beta}_{2SLS}$ and $\Cov \left[ \hat{\beta}_{2SLS} \right]$ are the estimated
coefficient and covariance matrix from 2SLS estimation, and
$\hat{\beta}_{3SLS}$ and $\Cov \left[ \hat{\beta}_{3SLS} \right]$ are the estimated
coefficients and covariance matrix from 3SLS estimation.
Under the null hypothesis this test statistic has a
$\chi^2$ distribution with degrees of freedom equal to the number of
estimated parameters.

### 2006-05-11 paper/tags/arne_diss_submit/systemfit_restrictions.tex
created tag for the revision that is included in Arnes dissertation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@667 591db2bd-e0a0-4d22-8c19-2192dbb7da31

It is common to perform hypothosis tests by imposing restrictions on
the parameter estimates.
There are two ways to impose linear parameter restrictions.
First, a matrix $T$ can be specified that
\begin{equation}
   \beta = T \cdot \beta^* \label{eq:T-restr} 
\end{equation}
where $\beta^*$ is a vector of restricted (linear independent) coefficients,
and $T$ is a matrix with the number of rows equal to the number of
unrestricted coefficients ($\beta$) and
the number of columns equal to the number of restricted coefficients
($\beta^*$).
$T$ can be used to map each unrestricted coefficient to one or more
restricted coefficients.

To impose these restrictions, the $X$ matrix is
(post-)multiplied by this $T$ matrix.
\begin{equation}
    X^* = X \cdot T
\end{equation}
Then, $X^*$ is substituted for $X$ and a standard estimation as described
in the previous section is done
(equations~\ref{eq:ols}--\ref{eq:3slsEViews}).
This results in the linear independent parameter estimates $\beta^*$ and
their covariance matrix.
The original parameters can be obtained by equation (\ref{eq:T-restr})
and the covariance matrix of the original parameters
can be obtained by:
\begin{equation}
   \Cov \left[ \widehat{\beta} \right] = T \cdot \Cov \left[ \widehat{\beta}^* \right] \cdot T'
\end{equation}

@@

The first way is less flexible than this latter one%
\footnote{
While restrictions like $\beta_1 = 2 \beta_2$ can be imposed by
both methods,
restrictions like $\beta_1 + \beta_2 = 4$ can be imposed only
by the second method.
}, 
but the first way is preferable if equality constraints for coefficients
across many equations of the system are imposed. 
Of course, these restrictions can be also imposed using
the latter method.
However, while the latter method increases the dimension of the 
matrices to be inverted during estimation, the first reduces it. 
Thus, in some cases the latter way leads to estimation problems
(e.g.\ (near) singularity of the matrices to be inverted),
while the first doesn't.

>>>>

It is common to perform hypothesis tests by imposing restrictions on
the parameter estimates.
There are two ways to impose linear parameter restrictions.
First, a matrix $T$ can be specified that
\begin{equation}
   \beta = T \cdot \beta^* \label{eq:T-restr} 
\end{equation}
where $\beta^*$ is a vector of restricted (linear independent) coefficients,
and $T$ is a matrix with the number of rows equal to the number of
unrestricted coefficients ($\beta$) and
the number of columns equal to the number of restricted coefficients
($\beta^*$).
$T$ can be used to map each unrestricted coefficient to one or more
restricted coefficients.

To impose these restrictions, the $X$ matrix is
(post-)\hspace{0pt}multiplied by this $T$ matrix.
\begin{equation}
    X^* = X \cdot T
\end{equation}
Then, $X^*$ is substituted for $X$ and a standard estimation as described
in the previous section is done
(equations~\ref{eq:ols}--\ref{eq:3slsEViews}).
This results in the linear independent parameter estimates $\beta^*$ and
their covariance matrix.
The original parameters can be obtained by equation (\ref{eq:T-restr})
and the covariance matrix of the original parameters
can be obtained by:
\begin{equation}
   \Cov \left[ \widehat{\beta} \right] = T \cdot \Cov \left[ \widehat{\beta}^* \right] \cdot T'
\end{equation}

@@

The first way is less flexible than this latter one%
\footnote{
While restrictions like $\beta_1 = 2 \beta_2$ can be imposed by
both methods,
restrictions like $\beta_1 + \beta_2 = 4$ can be imposed only
by the second method.
}, 
but the first way is preferable if equality constraints for coefficients
across many equations of the system are imposed. 
Of course, these restrictions can be also imposed using
the latter method.
However, while the latter method increases the dimension of the 
matrices to be inverted during estimation, the first reduces it. 
Thus, in some cases the latter way leads to estimation problems
(e.g.\ (near) singularity of the matrices to be inverted),
while the first does not.
